HO unification from object language to meta language

### 2024-02-24 paper/main.tex
wip

\noindent
Unfortunately this direct translation of rule \ref{clause:r3} uses the
predicate \coqIn{P} essentially as a first order term for the meta
language (its type is \elpiIn{tm}).
If we try to backchain the rule \ref{clause:r3} on the encoding of the goal \ref{goal:g} below:

@@

\noindent
we obtain an unsolvable unification problem \ref{problem:p}:
the two lists of terms have different lengths! The root cause is that
\ref{problem:p} is an higher order in DTT, but becomes
first order due to the encoding in the meta language.

@@

\begin{elpicode}
app[con"nfact", y, con"3"] = Pm y                     ~\customlabel{problem:pa}{(pa)}~
Pm = x\ app[con"nfact", x, con"3"]     % assignment for Pm
A = app[con"fin", con"7"]              % assignment for A
\end{elpicode}

\noindent
After unifying the head of rule \ref{clause:r3a} with the goal Elpi runs
the premise \elpiIn{link Pm A P} that is in charge of bringing the
assignment for \elpiIn{Pm} (that has type \elpiIn{tm -> tm})
back to the domain of Coq terms (the type \elpiIn{tm}):

@@

\noindent
This simple example is sufficient to show that the encoding we seek
is not trivial, since the solution for \elpiIn{P} above generates a
(Coq) $\beta$-redex in the second premise (the predicate
under the \elpiIn{pi w\ }\hspace{-0.8em}):

@@

\begin{elpicode}
app[lam A (a\ app[con"nfact", a, con"3"]), x] =
  app[con"nfact", N, NF]
\end{elpicode}

\noindent
The root cause of the problems we face is that the unification procedure
\Ue of the meta language is not aware of the equational theory of the object
logic \Uo, even if both theories include $\eta\beta$-conversion and admit most general
unifiers for problems in the pattern fragment~\cite{miller92jsc}.

@@

\section{Problem statement and alternative encodings} %%%%%%%%%%%%%%%%%%%%%%
\label{sec:encodings}

@@

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
$\sigma$, we write $\sigma t$ for the application of the substitution to
$t$ and we assume that the unification of our meta language is correct:

@@

Given 
$$
\C{s_1}{t_1}{l_1} \land \C{s_2}{t_2}{l_2}
$$
we define
$$
s_1 \Uo s_2 \mapsto \rho \stackrel{def}{=}
    t_1 \Ue t_2 \mapsto \sigma \land
    \D{\sigma}{l_1 + l_2}{\rho}
$$

We write $s \in \mathcal{L}_\lambda$ if all unif variables in
$s$ are applied to distinct bound variables.
%
\begin{gather}
\unifcorrect{\Eo}{\Uo}{s}\\
\unifcomplete{\Eo}{\Uo}{s}\\
% \forall \rho'\rho, 
  \rho s_1 \Eo \rho s_2 \Rightarrow
  \rho' \subseteq \rho \Rightarrow
  \rho's_i \in \mathcal{L}_\lambda \Rightarrow
  \rho' s_1 \Uo \rho' s_2
\end{gather}

the first one is that we are correct

the second one is that we are complete and mgu iff the problems are nice

the third one is still broken since we did not define how
to apply a fo subst to a ho term but it morally means
that any valid substitution for s1 s2 cannot break
the unification of terms t1 t2 obtained by compiling
s1 and s2 before knowing the subst.

These properties allow us to simulate a unification based
backward search on DTT by using \Ue, in a faithful way:
the trace of the logic program
performing the search not only gives the same result, but
also takes the same paths, that is it fails as early as possible.

\subsection{Alternative encodings and related work}

>>>>

\noindent
Unfortunately this translation of rule \ref{clause:r3} uses the
predicate \coqIn{P} as a first order term for the meta
language (its type is \elpiIn{tm}).
If we try to backchain the rule \ref{clause:r3} on the encoding of the goal
\ref{goal:g}:

@@

\noindent
we obtain an unsolvable unification problem \ref{problem:p}:
the two lists of terms have different lengths!
%The root cause is that
%\ref{problem:p} is an higher order in DTT, but becomes
%first order in the meta language due to the ``naive'' encoding.

@@

\begin{elpicode}
app[con"nfact", y, con"3"] = Pm y                     ~\customlabel{problem:pa}{(p')}~
Pm = x\ app[con"nfact", x, con"3"]     % assignment for Pm
A = app[con"fin", con"7"]              % assignment for A
\end{elpicode}

\noindent
After unifying the head of rule \ref{clause:r3a} with the goal Elpi runs
the premise \elpiIn{link Pm A P} that is in charge of bringing the
assignment for \elpiIn{Pm} (that has type \elpiIn{tm -> tm})
back to the domain of Coq terms:

@@

\noindent
This simple example is sufficient to show that the encoding we seek
is not trivial and does not only concern the head of rules, but the entire sequence
of unification problems that constitute the execution of a logic program.
In fact
the solution for \elpiIn{P} above generates a
(Coq) $\beta$-redex in the second premise (the predicate
under the \elpiIn{pi w\ }\hspace{-0.4em}):

@@

\begin{elpicode}
app[lam A (a\ app[con"nfact", a, con"3"]), x] =
app[con"nfact", N, NF]
\end{elpicode}
\noindent
~\\
The root cause of the problems we sketched in the running example
is that the unification procedure \Ue of the meta language is not aware
of the equational theory of the object logic, even if both theories
include $\eta\beta$-conversion and admit most general
unifiers for unification problems in the pattern fragment \llambda~\cite{miller92jsc}.

@@

\section{Problem statement} %%%%%%%%%%%%%%%%%%%%%%
\label{sec:encodings}

@@

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
$\sigma$, we write $\sigma t$ for the application of the substitution to
$t$, $\sigma \subseteq \sigma'$ when $\sigma$ is more general than $\sigma'$,
and we assume that the unification of our meta language is correct:

@@

Given 
$$
\C{s_1}{t_1}{l_1} \land \C{s_2}{t_2}{l_2}
$$
we define
$$
s_1 \Uo s_2 \mapsto \rho \stackrel{def}{=}
    t_1 \Ue t_2 \mapsto \sigma \land
    \D{\sigma}{l_1 + l_2}{\rho}
$$
Where $l_1 + l_2$ is the list concatenation of links.

We write $s \in \mathcal{L}_\lambda$ if all unif variables in
$s$ are applied to distinct bound variables.
%
\begin{gather}
\unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
\unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}\\
% \forall \rho'\rho, 
  \rho s_1 \Eo \rho s_2 \Rightarrow
  \rho' \subseteq \rho \Rightarrow
  \rho's_i \in \mathcal{L}_\lambda \Rightarrow
  \rho' s_1 \Uo \rho' s_2 \label{prop:simulation}
\end{gather}

Properties \ref{prop:correct} and \ref{prop:complete} state, respectively, that
in \llambda the implementation of \Uo is correct, complete and returns the most
general unifier.

Property \ref{prop:simulation} states that \Uo is resilient to problems outside
\llambda if a third party provides a (partial) solution for its problem. Since
we are interested in using \Uo in a proof search procedure, made of a sequence
of unification problems, not necessarily in \llambda. In practice it is often
the case the order in which these problems are stated matters. A Typical example
is the following problem

\begin{elpicode}
app [F, con"a"] = app[F, con"a", con"a"]
\end{elpicode}

\noindent
preceded by

\begin{elpicode}
F = lam x\app[const f,x,x]
\end{elpicode}

\noindent
becomes solvable in DTT trivially, since the term is ground (hence in \llambda),
but is one substitutes F in the LHS does not find, structurally, the RHS
hence \Ue would fail (since \Ue does not know about the $\beta$ rule of DTT).
Our compiler takes care of making property \ref{prop:simulation} hold, see
section XXX.

Property \ref{prop:complete} is also relevant to use \Uo for logic programming.
In particular we want failures to occur as early as possible, so want
the decomp phase to take place immediately after \Ue, and fail if need be.
This becomes particularly important since compile may introduce two
ho variables for the same fo one, leaving the task of unifying the solutions
to decomp.

\section{Alternative encodings and related work}
